---
title: "Co-occurrence Network Analysis — SparCC Bootstrap"
author: "Vy Nguyen"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
params:
  run_name:       "HGM_study"
  output_dir:     "output"
  sparcc_iter:    50      # SparCC iterations
  inner_iter:     20      # SparCC inner iterations
  bootstrap_r:    500     # Bootstrap replicates (500 for publication)
  cor_cutoff:     0.3     # Minimum absolute correlation to retain
  pval_cutoff:    0.05    # FDR threshold
  stab_cutoff:    0.8     # Minimum bootstrap stability
  min_prevalence: 0.3     # Minimum prevalence for species inclusion (30%)
---

> **Note**: Networks are constructed separately per enterotype group, resulting
> in small effective sample sizes. Results should be interpreted as exploratory.
> Claims about hub taxa and network properties are descriptive only and require
> independent validation in larger cohorts.

```{r setup, message=FALSE, warning=FALSE}
library(SpiecEasi)
library(igraph)
library(ggraph)
library(ggplot2)
library(dplyr)
library(writexl)
library(readxl)
library(tibble)
library(pheatmap)
library(RColorBrewer)
library(gridExtra)
library(here)

run_name       <- params$run_name
output_dir     <- params$output_dir
SPARCC_ITER    <- params$sparcc_iter
SPARCC_INNER_ITER <- params$inner_iter
BOOTSTRAP_R    <- params$bootstrap_r
COR_CUTOFF     <- params$cor_cutoff
PVAL_CUTOFF    <- params$pval_cutoff
STAB_CUTOFF    <- params$stab_cutoff
MIN_PREV       <- params$min_prevalence
prefix         <- paste0(run_name, "_")

dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Limit BLAS/OpenMP threads to avoid parallelism conflicts with sparccboot
Sys.setenv(
  OPENBLAS_NUM_THREADS = "1",
  OMP_NUM_THREADS      = "1",
  MKL_NUM_THREADS      = "1",
  VECLIB_MAXIMUM_THREADS = "1",
  NUMEXPR_NUM_THREADS  = "1"
)
if (requireNamespace("RhpcBLASctl", quietly = TRUE)) {
  RhpcBLASctl::blas_set_num_threads(1)
  RhpcBLASctl::omp_set_num_threads(1)
}

cat(sprintf("Parameters: Iter=%d, Bootstrap=%d, Cor_Th=%.2f, Pval_Th=%.2f, Stab_Th=%.2f\n",
            SPARCC_ITER, BOOTSTRAP_R, COR_CUTOFF, PVAL_CUTOFF, STAB_CUTOFF))
```

## 1. Load Data

```{r load-data}
# Raw count matrix (required for SparCC)
otu_raw_data   <- read_excel(here("data", "species_abundance.xlsx"),
                              sheet = "species_counts_transposed")
otu_matrix_raw <- as.matrix(otu_raw_data[, -1])
rownames(otu_matrix_raw) <- otu_raw_data[[1]]
otu_matrix_raw <- round(otu_matrix_raw)
storage.mode(otu_matrix_raw) <- "integer"

# Metadata
metadata <- read_excel(here("data", "metadata.xlsx"), sheet = "metadata")
metadata <- as.data.frame(metadata)
rownames(metadata) <- metadata$SampleID

# Align samples
common_samples <- intersect(colnames(otu_matrix_raw), rownames(metadata))
otu_raw_clean  <- otu_matrix_raw[, common_samples]
meta_clean     <- metadata[common_samples, ]

cat("Samples:", ncol(otu_raw_clean),
    "| Species:", nrow(otu_raw_clean), "\n")
cat("Enterotype distribution:\n")
print(table(meta_clean$Enterotype))
```

## 2. Prepare SparCC Input Matrix

```{r prepare-matrix}
# Transpose raw count matrix to samples x species
otu_mat <- t(otu_raw_clean)
storage.mode(otu_mat) <- "integer"

# Prevalence filter: retain species present in >= 30% of samples
min_prev      <- MIN_PREV * nrow(otu_mat)
keep_taxa     <- colSums(otu_mat > 0) >= min_prev
otu_mat_clean <- otu_mat[, keep_taxa]

cat("After prevalence filter:", ncol(otu_mat_clean), "species retained\n")
```

## 4. SparCC Analysis Function

```{r sparcc-function}
analyze_sparcc <- function(data_mat, label) {

  message("\n>>> Analyzing: ", label,
          " | Dimensions: ", paste(dim(data_mat), collapse = "x"))

  if (ncol(data_mat) < 3) {
    message("   Skipped: fewer than 3 species.")
    return(NULL)
  }

  # Use available cores minus 2, minimum 1
  n_cores <- max(1, parallel::detectCores() - 2)

  set.seed(123)

  # Run SparCC bootstrap
  res_boot <- tryCatch(
    sparccboot(data = data_mat,
               R    = BOOTSTRAP_R,
               ncpus = n_cores,
               sparcc.params = list(iter       = SPARCC_ITER,
                                    inner_iter  = SPARCC_INNER_ITER,
                                    th          = 0.1)),
    error = function(e) {
      message("   SparCC error: ", e$message)
      NULL
    }
  )
  if (is.null(res_boot)) return(NULL)

  # Extract p-values and correlations
  res_pval   <- pval.sparccboot(res_boot)
  cor_matrix <- res_pval$cors
  p_matrix   <- res_pval$pvals

  # Reshape if output is unexpectedly flat
  if (is.null(dim(cor_matrix))) {
    n_taxa     <- ncol(data_mat)
    cor_matrix <- matrix(cor_matrix, nrow = n_taxa, ncol = n_taxa)
    p_matrix   <- matrix(p_matrix,   nrow = n_taxa, ncol = n_taxa)
    colnames(cor_matrix) <- rownames(cor_matrix) <- colnames(data_mat)
    colnames(p_matrix)   <- rownames(p_matrix)   <- colnames(data_mat)
  }

  # Bootstrap stability: fraction of bootstrap replicates matching sign of observed correlation
  cor_vec <- cor_matrix[lower.tri(cor_matrix)]
  if (!is.null(res_boot$t) && ncol(res_boot$t) == length(cor_vec)) {
    stability_vec <- sapply(seq_along(cor_vec), function(idx) {
      boot_vals <- res_boot$t[, idx]
      mean(sign(boot_vals) == sign(cor_vec[idx]), na.rm = TRUE)
    })
  } else {
    message("   Warning: stability calculation skipped (dimension mismatch).")
    stability_vec <- rep(NA, length(cor_vec))
  }

  # Build edge list from upper triangle
  taxa_names <- colnames(data_mat)
  edges      <- data.frame()
  k          <- 1

  for (i in seq_len(length(taxa_names) - 1)) {
    for (j in seq(i + 1, length(taxa_names))) {
      r_val <- cor_matrix[i, j]
      p_raw <- p_matrix[i, j]
      stab  <- if (k <= length(stability_vec)) stability_vec[k] else NA
      k     <- k + 1

      # Pseudo-count correction for p-values of zero
      p_corrected <- max(p_raw, 1 / (BOOTSTRAP_R + 1))

      if (!is.na(r_val) && abs(r_val) >= COR_CUTOFF) {
        edges <- rbind(edges, data.frame(
          from        = taxa_names[i],
          to          = taxa_names[j],
          weight      = r_val,
          p_raw       = p_corrected,
          stability   = stab,
          group       = label,
          stringsAsFactors = FALSE
        ))
      }
    }
  }

  if (nrow(edges) == 0) {
    message("   No edges above |r| threshold.")
    return(NULL)
  }

  # FDR correction
  edges$p_adj      <- p.adjust(edges$p_raw, method = "BH")
  edges_filtered   <- edges %>%
    filter(p_adj < PVAL_CUTOFF) %>%
    mutate(interaction = ifelse(weight > 0, "Positive", "Negative"))

  message("   Significant edges (FDR < ", PVAL_CUTOFF, "): ", nrow(edges_filtered))
  return(edges_filtered)
}
```

## 5. Network Plotting Function

```{r network-function}
plot_network <- function(edges_df, label) {

  g <- graph_from_data_frame(edges_df, directed = FALSE)
  if (vcount(g) > 0) g <- delete_vertices(g, which(degree(g) == 0))
  if (vcount(g) == 0) return(NULL)

  # Edge distance weight for path calculations
  E(g)$distance <- 1 / abs(E(g)$weight)

  # Node centrality metrics
  V(g)$degree      <- degree(g)
  V(g)$betweenness <- betweenness(g, normalized = TRUE, weights = NA)
  V(g)$closeness   <- closeness(g, normalized = TRUE, weights = NA)
  V(g)$eigen       <- eigen_centrality(g, weights = abs(E(g)$weight))$vector
  V(g)$clustering_coeff <- transitivity(g, type = "local", vids = V(g))

  # Community detection on positive edges only
  g_pos  <- subgraph.edges(g, E(g)[weight > 0], delete.vertices = FALSE)
  comm   <- cluster_louvain(g_pos, weights = E(g_pos)$weight)
  V(g)$community <- as.factor(membership(comm))

  # Hub detection: top 20th percentile for both degree and betweenness
  V(g)$is_hub <- V(g)$degree      >= quantile(V(g)$degree, 0.80) &
                 V(g)$betweenness >= quantile(V(g)$betweenness, 0.80)

  # Global network metrics
  num_pos  <- sum(E(g)$weight > 0)
  num_neg  <- sum(E(g)$weight < 0)

  metrics_df <- data.frame(
    Network           = label,
    Nodes             = vcount(g),
    Edges             = ecount(g),
    Pos_Edges         = num_pos,
    Neg_Edges         = num_neg,
    Pos_Neg_Ratio     = round(ifelse(num_neg > 0, num_pos / num_neg, num_pos), 2),
    Density           = round(edge_density(g), 4),
    Global_Clustering = round(transitivity(g, type = "global"), 4),
    Avg_Clustering    = round(mean(V(g)$clustering_coeff, na.rm = TRUE), 4),
    Modularity        = round(modularity(comm), 4),
    Avg_Degree        = round(mean(V(g)$degree), 3),
    Avg_Path_Len      = round(mean_distance(g, directed = FALSE, weights = NA), 3)
  )

  # Node-level table
  nodes_df <- data.frame(
    Species          = V(g)$name,
    Degree           = V(g)$degree,
    Betweenness      = round(V(g)$betweenness, 4),
    Closeness        = round(V(g)$closeness, 4),
    Eigenvector      = round(V(g)$eigen, 4),
    Clustering_Coeff = round(V(g)$clustering_coeff, 4),
    Community        = V(g)$community,
    Is_Hub           = V(g)$is_hub,
    Network          = label
  ) %>% arrange(desc(Betweenness))

  # Plot
  set.seed(999)
  lay <- layout_with_graphopt(g, charge = 0.05, spring.length = 20)
  V(g)$label_show <- ifelse(V(g)$is_hub, V(g)$name, NA)

  p <- ggraph(g, layout = lay) +
    geom_edge_link(aes(color = interaction, width = abs(weight)), alpha = 0.5) +
    scale_edge_color_manual(values = c("Positive" = "#3498DB",
                                       "Negative" = "#E74C3C")) +
    scale_edge_width(range = c(0.3, 1.5), guide = "none") +
    geom_node_point(aes(size = degree, fill = community),
                    shape = 21, color = "white", stroke = 0.5) +
    scale_size(range = c(3, 10)) +
    scale_fill_brewer(palette = "Set3") +
    geom_node_text(aes(label = label_show),
                   repel = TRUE, size = 3, fontface = "bold") +
    theme_void() +
    labs(title    = paste("Co-occurrence Network —", label),
         subtitle = sprintf("Nodes: %d | Edges: %d (+%d/-%d) | Clustering: %.3f",
                            vcount(g), ecount(g), num_pos, num_neg,
                            transitivity(g, type = "global")),
         caption  = "Labelled nodes = hub taxa (top 20th percentile: degree + betweenness)")

  return(list(metrics = metrics_df, nodes = nodes_df, plot = p))
}
```

## 6. Run Analysis

```{r run-global}
all_edges   <- list()
all_metrics <- list()

# Global network (all samples)
e_glob <- analyze_sparcc(otu_mat_clean, "Global")
if (!is.null(e_glob)) {
  all_edges[["Edges_Global"]] <- e_glob
  res <- plot_network(e_glob, "Global")
  if (!is.null(res)) {
    all_metrics[["Global"]]     <- res$metrics
    all_edges[["Nodes_Global"]] <- res$nodes
    print(res$plot)
    ggsave(file.path(output_dir, paste0(prefix, "Global.png")),
           res$plot, width = 12, height = 10, dpi = 300, bg = "white")
  }
}
```

```{r run-enterotypes}
# Per-enterotype networks
for (et in unique(meta_clean$Enterotype)) {
  samps <- intersect(rownames(meta_clean)[meta_clean$Enterotype == et],
                     rownames(otu_mat_clean))

  if (length(samps) < 5) {
    cat("Skipping", et, "— fewer than 5 samples\n")
    next
  }

  sub_mat <- otu_mat_clean[samps, , drop = FALSE]
  sub_mat <- sub_mat[, colSums(sub_mat) > 0, drop = FALSE]

  e_et <- analyze_sparcc(sub_mat, et)
  if (!is.null(e_et)) {
    all_edges[[paste0("Edges_", et)]] <- e_et
    res <- plot_network(e_et, et)
    if (!is.null(res)) {
      all_metrics[[et]]                  <- res$metrics
      all_edges[[paste0("Nodes_", et)]]  <- res$nodes
      print(res$plot)
      ggsave(file.path(output_dir, paste0(prefix, "Network_", et, ".png")),
             res$plot, width = 12, height = 10, dpi = 300, bg = "white")
    }
  }
}
```

## 7. Export Results

```{r export}
if (length(all_metrics) > 0) {
  network_summary <- do.call(rbind, all_metrics)
  all_edges[["Network_Summary"]] <- network_summary

  cat("\n=== Network Summary ===\n")
  print(network_summary)

  write_xlsx(all_edges,
             file.path(output_dir, paste0(prefix, "SparCC_Results_FDR.xlsx")))
  cat("Results exported.\n")
} else {
  cat("No significant networks found.\n")
}
```

## Session Info

```{r session-info}
sessionInfo()
```
